---
title: "DATA PROCESSING"
authors: "Marco D'Agostino, Théo Degand, Lei Ding"
toc: true
format: 
  html:
    output-file: "index.html"
    html-math-method: mathjax
    code-fold: false
    # code-summary: "Show code"
    code-tools: true
    code-overflow: wrap
    self-contained: false
    theme: zephyr
editor: source
---

## INTRODUCTION & SET UP

This notebook presents simple codes to load, process and clean time-series data about leaf water potential ( $LWP$ ) , soil water potential ($SWP$) and weight ($W$) for transpiration rate ($T_r$).

The repository contains the following folders :

-   **data/** :

    -   **raw/** contains one folder per variable. Inside each subfolder are stored raw files directly collected from sensors. Typically, each file corresponds to one plant.

    -   **clean/** contains cleaned tables (in .csv format) for metadata and for each variable.

-   **plots/** : contains plots and figures generated by this notebook.

-   **src/** contains custom functions used to process, compile and clean the raw data.

### Loading libraries and defining paths

```{r, results='hide', message=FALSE, warning=FALSE}
# rm(list=ls())                                               # clean workspace [comment when Rendering]
# setwd(dirname(rstudioapi::getActiveDocumentContext()$path)) # Set working directory to the current file [comment when Rendering]

# LOAD LIBRARIES
library(SciViews)
library(ggplot2)
library(pander)
library(dplyr)
library(tidyverse)
library(data.table)
library(anytime)
library(ggthemes)

# LOAD UTILITARY FUNCTIONS
source("src/Data_Preparation.R")  # Contains functions to extract data from raw files
source("src/LWP_algorithm.R")     # Contains the code of the LWP cleaning algorithm

# SET DIRECTORIES
data_path_clean <- paste0("data/clean/") # Set path to Clean folder
data_path_raw   <- paste0("data/raw/")   # Set path to row folder
plot_path       <- paste0("plots/")      # Set path to plot folder

path_LWP        <- paste0(data_path_raw, "LWP/")
path_SWP        <- paste0(data_path_raw, "SWP/")
path_weight     <- paste0(data_path_raw, "weight/")
```

```{r,}
my_theme <- theme_classic() +
  theme(
    text            = element_text(size = 11),
    axis.title      = element_text(size = 12),
    axis.text       = element_text(size = 11),
    axis.text.x     = element_text(size = 8, angle = 90, hjust = 1, vjust = 0.5),
    plot.title      = element_text(size = 12, hjust = 0.5),
    strip.background = element_rect(colour = "white", fill = "white"),
    strip.text = element_text(size = 12), # Increase size and make bold
  )
```

### Load Metadata

Load metadata about the experiment.

Here we designed a simple experiment as example :

-   4 plants (`Plant_ID` column)

-   2 different treatments (`Treatment` column) :

    -   well watered (*WW*)

    -   water deficit (*WD* )

-   Leaf water potential ($LWP$) was recorded with psychrometers (`PSY` column). Raw files are stored in `data/raw/LWP/`.

-   Soil water potential ($SWP$) was recorded with TEROS 21 (`Teros` column). Raw files are stored in `data/raw/SWP/`.

-   Weight ($W$) was recorded with scales (`Scale` column). Raw files are stored in `data/raw/weight/`.

```{r}
Plants_ID <- read.csv(paste0(data_path_clean, "Plants_ID.csv"))
pander(Plants_ID)
```

## 1) LOAD LEAF WATER POTENTIAL DATA

### Extract from raw files

$LWP$ data from psychrometers are stored in separated `.csv` files.

We use the custom function `load_LWP()` to load and process each csv file in the provided path.

The result is a single table containing Date, Time, LWP value and the ID of plants.

The single file is written as `LWP_raw.csv` in the raw folder.

```{r, eval=FALSE}
# Compile all PSY files into one dataframe
LWP_raw <- load_LWP(path = path_LWP)

# Change the PSY id by Plant_ID
LWP_raw <- left_join(LWP_raw, 
                     Plants_ID[c("Plant_ID", "PSY")], 
                     by = "PSY") %>%
  select(-c("PSY"))

pander(head(LWP_raw))

# Write to raw folder
write.csv(x = LWP_raw, file = paste0(data_path_raw, "LWP_raw.csv"), row.names = F)
```

### Load compiled raw file

Once the raw files are compiled, we can simply load the processed csv file.

NB : each time we load the raw.csv files, we need to format correctly the DateTime column.

```{r, message=FALSE, warning=FALSE}
# Load from Raw folder
LWP_raw <- read.csv(paste0(data_path_raw, "LWP_raw.csv")) %>% 
  mutate(DateTime = anytime(DateTime)) %>% 
  mutate(DateTime = as.POSIXct(DateTime, format = "%Y-%m-%d %H:%M:%S"))
```

Simple plot :

```{r, message=FALSE, warning=FALSE}
p1 <- ggplot(left_join(LWP_raw, Plants_ID, by = "Plant_ID")) + 
  
  # Display data as line
  geom_line(aes(x = DateTime, 
                y = LWP, 
                color = Treatment,
                group = Plant_ID)) +
  
  # Split into facets per Plant_ID
  facet_wrap(~Plant_ID, 
             scales = "free",
             ncol = 2) +
  
  # Setup x axis
  scale_x_datetime(date_breaks = "1 day",
                   date_labels = "%d/%m",
                   name = "Date") + 
  
  # Setup y axis
  ylab("Leaf Water Potantial [MPa]") +
  ylim(-1.5, 0.5) +
  
  # Add custom theme
  my_theme
  
# ========================= # 
# Show plot and save to file
p1
ggsave(filename = "LWP_raw.png", plot = p1, device = "png", path = plot_path, width = 9, height = 4)
```

### Apply cleaning algorithm and store to folder.

We see that data is messy, so we implemented an algorithm to clean LWP data. It consists of creating buffers per day, and for each buffer we run a bunch of tests to check if we want to keep the buffer of not.

-   `check.na` evaluates how many NAN are in the buffer. If there is more NAN than the `nan.th` proportion, the buffer is discarded.

-   `check.zero` evaluates how many zeros are in the buffer. If there is more zeros than the `zero.th` proportion, the buffer is discarded.

-   `check.cycles` evaluates if LWP at night is lesser than LWP at day. If the mean LWP at day is higher that the mean LWP at night multiplied by `cycle.alpha`, the buffer is discarded.

-   `check.slope` evaluates the mean slope of the buffer via a linear model. If the absolute value of the slope is higher than `slope.th`, the buffer is discarded.

-   `check.level` evaluates the statistical mean of the buffer compared to the mean of the other buffers. If $\mu_{LWP, i} < \mu_{LWP} - \alpha \cdot \sigma_{LWP}$ or $\mu_{LWP, i} > \mu_{LWP} + \alpha \cdot \sigma_{LWP}$ (with $\mu_{LWP}$ is the mean LWP for all buffers, $\mu_{LWP,i}$ is the mean LWP of the ith buffer, $\alpha$ is the unput multiplicator and $\sigma_{LWP}$ is the standard deviation of LWP), the buffer is discarded.

Finally, the clean dataframe is written to storage.

```{r, message=FALSE, warning=FALSE}
LWP_clean <- clean.LWP(LWP_data = LWP_raw,
                       check.na     = T,
                       check.zero   = T, 
                       check.cycles = T, 
                       check.slope  = F, 
                       check.level  = F, 
                       nan.th       = 0.5, 
                       zero.th      = 0.5, 
                       cycle.alpha  = 0.5, 
                       slope.th     = 0.000002, 
                       level.alpha  = 2)

LWP_clean <- LWP_clean[c("Date", "Time", "DateTime", "LWP", "Plant_ID")]

# Additionnal cleaning
LWP_clean <- LWP_clean %>%
  filter(LWP > -3)

# Write to storage
write.csv(x = LWP_clean, file = paste0(data_path_clean, "LWP_clean.csv"), row.names = F)
```

### Load clean file

```{r, message=FALSE, warning=FALSE}
LWP_clean <- read.csv(paste0(data_path_clean, "LWP_clean.csv")) %>% 
  mutate(DateTime = anytime(DateTime)) %>% 
  mutate(DateTime = as.POSIXct(DateTime, format = "%Y-%m-%d %H:%M:%S"))

# Add NA so that there is gaps in geom_line
temp <- expand.grid(DateTime = unique(LWP_clean$DateTime),
                    Plant_ID = unique(LWP_clean$Plant_ID)
                    )

LWP_clean <- left_join(temp, LWP_clean, by = c("Plant_ID", "DateTime"))
```

Simple plot :

```{r, message=FALSE, warning=FALSE}
LWP_plot <- ggplot(left_join(LWP_clean, Plants_ID, by = "Plant_ID")) + 
  
  # Display data as line
  geom_line(aes(x = DateTime, 
                y = LWP, 
                color = Treatment,
                group = Plant_ID)) +
  
  # Split into facets per Plant_ID
  facet_wrap(~Plant_ID, 
             scales = "free",
             ncol = 2) +
  
  # Setup x axis
  scale_x_datetime(date_breaks = "1 day",
                   date_labels = "%d/%m",
                   name = "Date") + 
  
  # Setup y axis
  ylab("Leaf Water Potantial [MPa]") +
  ylim(-1.5, 0.5) +
  
  # Add custom theme
  my_theme

# ========================= # 
# Show plot and save to file
LWP_plot
ggsave(filename = "LWP_clean.png", plot = LWP_plot, device = "png", path = plot_path, width = 9, height = 4)
```

We can make a quick graph of the comparison between Raw and Clean LWP

```{r, message=FALSE, warning=FALSE}
p <- ggplot() +
  geom_line(data = LWP_raw, aes(x = DateTime, y = LWP), color = "red") +
  geom_line(data = LWP_clean, aes(x = DateTime, y = LWP), color = "green") +
  # ylim(-3, 0.5) +
  facet_wrap(~Plant_ID, scale = "free") +
  my_theme

# ========================= # 
# Show plot and save to file
p
ggsave(filename = "LWP_comparison.png", 
       plot = p, device = "png", path = plot_path, width = 9, height = 4)
```

## 2) LOAD SOIL WATER POTENTIAL DATA

### Extract from sensor files and compile into a single raw file

SWP data si stored as a unique `.DAT` file. \
`load_SWP` function load the file and convert it to dataframe. \
The pre-processed file can then be written to `data/raw/` folder.

```{r, eval=FALSE, message=FALSE, warning=FALSE}
# Extract from DAT files
SWP_raw <- load_SWP(path_SWP = path_SWP)

# Replace Teros_ID by Plant_ID and set DateTime as POSIXct
SWP_raw <- left_join(SWP_raw, Plants_ID[c("Plant_ID", "Teros")], by = c("Teros_ID" = "Teros")) %>% 
    select(c("DateTime", "SWP", "Plant_ID")) %>% 
    mutate(DateTime = as.POSIXct(DateTime, format = "%d/%m/%Y %H:%M"))

# Write to disk
write.csv(x = SWP_raw, file = paste0(data_path_raw, "SWP_raw.csv"), row.names = F)
```

### Load raw file, clean it a bit and write to a clean file

Here we load the pre-processed SWP dataframe. Statistical or manual corrections can be done at will. \
The final version of SWP data can then be written to `data/clean/` folder.

```{r, message=FALSE, warning=FALSE}
# Load SWP_raw
SWP_raw <- na.omit(read.csv(paste0(data_path_raw, "SWP_raw.csv"))) %>% 
  mutate(DateTime = anytime(DateTime)) %>% 
  mutate(DateTime = as.POSIXct(DateTime, format = "%Y-%m-%d %H:%M:%S"))

# Filter data so DateTime match with LWP
SWP_clean <- SWP_raw %>% 
  filter(DateTime >= min(LWP_clean$DateTime) & DateTime <= max(LWP_clean$DateTime)) %>% 
  filter(DateTime > as.POSIXct("2024-10-25"))

# Write to disk
write.csv(x = SWP_clean, file = paste0(data_path_clean, "SWP_clean.csv"), row.names = F)
```

Simple plot :

```{r, message=FALSE, warning=FALSE}
SWP_plot <- ggplot(left_join(SWP_clean, Plants_ID, by = "Plant_ID")) +
  # Display data as line
  geom_line(aes(x = DateTime, 
                y = SWP, 
                color = Treatment,
                group = Plant_ID)) +
  
  # Split into facets per Plant_ID
  facet_wrap(~Plant_ID, 
             scales = "free",
             ncol = 2) +
  
  # Setup x axis
  scale_x_datetime(date_breaks = "1 day",
                   date_labels = "%d/%m",
                   name = "Date") + 
  
  # Setup y axis
  ylab("Soil Water Potantial [MPa]") +
  ylim(-850, 0.5) +
  
  # Add custom theme
  my_theme

# =========================================== #
SWP_plot
ggsave(filename = "SWP_clean.png", plot = SWP_plot, device = "png", path = plot_path, width = 9, height = 4)
```

## 3) EXTRACT WEIGHT DATA

### Extract from sensor output files and compile into one single raw file

Weight data is stored as individual `.csv` files in `data/raw/weight/` folder. \
Similarly to LWP and SWP, the function `load_W` function extracts 
individual files and creates a single table. \
The table is store in `data/raw/`folder.

```{r, eval=FALSE, message=FALSE, warning=FALSE}
# Load raw data
W_raw <- load_W(path = path_weight, skip = 1) %>% 
  inner_join(Plants_ID[c("Plant_ID", "Scale")], by = "Scale") %>% 
  mutate(Date = as.Date(Date, format = "%d/%m/%Y")) %>%
  select(-c("Scale"))

# Write to disk
write.csv(x = W_raw, file = paste0(data_path_raw, "W_raw.csv"), row.names = F)
```

### Load raw file, clean a bit and write clean file

Here we load the pre-processed Weight table, apply corrections and filtering at will, then write cleaned data to `data/clean/` folder.

```{r, message=FALSE, warning=FALSE}
# Load raw file
W_raw <- unique(read.csv(paste0(data_path_raw, "W_raw.csv"))) %>% 
  mutate(DateTime = anytime(DateTime)) %>% 
  mutate(DateTime = as.POSIXct(DateTime, format = "%Y-%m-%d %H:%M:%S"))

# Cleaning
W_clean <- W_raw %>% 
  filter(DateTime >= as.POSIXct("2024-10-24 11:00:00"))

# Write to disk
write.csv(x = W_clean, file = paste0(data_path_clean, "W_clean.csv"), row.names = F)
```

Simple plot :

```{r, message=FALSE, warning=FALSE}
W_plot <- ggplot(W_clean %>% left_join(Plants_ID, by = "Plant_ID")) +
  
  # Display data as line
  geom_line(aes(x = DateTime, 
                y = Weight, 
                color = Treatment,
                group = Plant_ID)) +
  
  # Split into facets per Plant_ID
  facet_wrap(~Plant_ID, 
             scales = "free",
             ncol = 2) +
  
  # Setup x axis
  scale_x_datetime(date_breaks = "1 day",
                   date_labels = "%d/%m",
                   name = "Date") + 
  
  # Setup y axis
  ylab("Weight [g]") +
  # ylim(-1.5, 0.5) +
  
  # Add custom theme
  my_theme

W_plot
ggsave(filename = "W_clean.png", plot = W_plot, device = "png", path = plot_path, width = 9, height = 4)
```

## 4) Compute Transpiration rate

The function `compute_Tr` computes transpiration rate $T_r$ from Weight time series using the formula 

$$ 
T_{r, t} [g \cdot h^{-1}] = (W_{t-1} - W_{t}) / \Delta t  
$$

Pre-processed transpiration rate data is stored in `data/raw/` folder.

```{r, message=FALSE, warning=FALSE}
# Load clean Weight file
W_clean <- unique(read.csv(paste0(data_path_clean, "W_clean.csv"))) %>% 
  mutate(DateTime = anytime(DateTime)) %>% 
  mutate(DateTime = as.POSIXct(DateTime, format = "%Y-%m-%d %H:%M:%S"))

# Compute Tr using custom function
Tr_raw <- compute_Tr(W = W_clean, th = 20, DT.format = "%Y-%m-%d %H:%M:%S")

# Write to disk
write.csv(x = Tr_raw, file = paste0(data_path_raw, "Tr_raw.csv"), row.names = F)
```

### Load Tr raw from file

```{r, message=FALSE, warning=FALSE}
# Read Tr raw fil from folder
Tr_raw <- unique(read.csv(paste0(data_path_raw, "Tr_raw.csv"))) %>% 
  mutate(DateTime = anytime(DateTime)) %>% 
  mutate(DateTime = as.POSIXct(DateTime, format = "%Y-%m-%d %H:%M:%S"))
```

Simple plot

```{r}
Tr_plot <- ggplot(left_join(Tr_raw, Plants_ID, by = "Plant_ID")) +
  # Display data as line
  geom_line(aes(x = DateTime, 
                y = Tr, 
                color = Treatment,
                group = Plant_ID)) +
  
  # Split into facets per Plant_ID
  facet_wrap(~Plant_ID, 
             scales = "free",
             ncol = 2) +
  
  # Setup x axis
  scale_x_datetime(date_breaks = "1 day",
                   date_labels = "%d/%m",
                   name = "Date") + 
  
  # Setup y axis
  ylab("Transpiration rate [g/h]") +
  ylim(-1, 15) +
  
  # Add custom theme
  my_theme

Tr_plot

ggsave(filename = "Tr_raw.png", 
       plot = Tr_plot, device = "png", path = plot_path, width = 9, height = 4)
```

## Make one plot with the 3 variables

Create a gigantic file with every time series measurement.

For that, we create temporary table by gathering variables into one key column (`Measure`) and one value column (`Value`).

Then we bind them by row. For better visualization, we create a new attribute `Plant_name` that combine `Plant_ID` and `Treatment`.

```{r}
LWP_temp <- LWP_clean[c("Plant_ID", "DateTime", "LWP")] %>% 
  gather(LWP, key = "Measure", value = "Value") 

SWP_temp <- SWP_clean[c("Plant_ID", "DateTime", "SWP")] %>% 
  gather(SWP, key = "Measure", value = "Value") %>% 
  mutate(Value = Value/1000)

W_temp   <- W_clean[c("Plant_ID", "DateTime", "Weight")] %>% 
  gather(Weight, key = "Measure", value = "Value") %>% 
  mutate(Value = Value/1500)

T_temp   <- Tr_raw[c("Plant_ID", "DateTime", "Tr")] %>% 
  gather(Tr, key = "Measure", value = "Value") %>% 
  mutate(Value = Value/10)

TS_full <- rbind(LWP_temp, SWP_temp) %>% 
  rbind(T_temp) %>% 
  left_join(Plants_ID, by = "Plant_ID") %>% 
  mutate(Plant_name = paste0("Plant ", Plant_ID, " | ", Treatment))
```

Generate the plot

```{r}
full_plot <- ggplot(TS_full) +
  
  # Display the data as a line
  geom_line(aes(x = DateTime, y = Value, color = Measure)) +
  
  # Choose manual colors
  scale_color_manual(values = c("darkgreen", "orange", "blue")) +
  
  # Set x limits and define the format of breaks
  scale_x_datetime(# date_breaks = "1 day",
                   date_labels = "%d/%m",
                   name = "Date") + 
  
  # Set y limits and add a secondary axis
  scale_y_continuous(limits = c(-2, 1.5),
                     bquote(expr = psi[Soil/Leaf]~"[MPa]"), 
                     sec.axis = sec_axis(~.*10, name = bquote(T[act]~"["*gh^-1*"]"))
                     ) +
  
  # Split screen by Plant_name
  facet_wrap(~Plant_name, 
             scales = "free",
             ncol = 2) +
  
  # Premade theme
  my_theme +
  
  # Change some specificities of the theme
  theme(legend.position = "bottom") 

full_plot
ggsave(filename = "Full_plot.svg", plot = full_plot, device = "svg", path = plot_path, width = 7, height = 5)
ggsave(filename = "Full_plot.png", plot = full_plot, device = "png", path = plot_path, width = 7, height = 5)
```
