---
title: "Data Processing"
authors: "Marco D'Agostino, Lei Ding"
toc: true
format: 
  html:
    html-math-method: mathjax
    code-fold: false
    # code-summary: "Show code"
    code-tools: true
    code-overflow: wrap
    self-contained: false
    theme: zephyr
editor: source
---

## 0) Set up the notebook

### Loading libraries and defining paths

```{r, results='hide', message=FALSE, warning=FALSE}
rm(list=ls())                                               # clean workspace
setwd(dirname(rstudioapi::getActiveDocumentContext()$path)) # Set working directory to the current file

# LOAD LIBRARIES
library(SciViews)
library(ggplot2)
library(pander)
library(dplyr)
library(tidyverse)
library(data.table)
library(anytime)
library(ggthemes)

# LOAD UTILITARY FUNCTIONS
source("../src/Data_Preparation.R")  # Contains functions to extract data from raw files
source("../src/LWP_algorithm.R")     # Contains the code of the LWP cleaning algorithm

# SET DIRECTORIES
ExperimentID    <- "Protocol_maize"

data_path_clean <- paste0("../data/experiments/", ExperimentID, "/clean/") # Set path to Clean folder
data_path_raw   <- paste0("../data/experiments/", ExperimentID, "/raw/")   # Set path to row folder
plot_path       <- paste0("../plots/", ExperimentID,"/")                   # Set path to plot folder

path_LWP        <- paste0(data_path_raw, "LWP/")
path_SWP        <- paste0(data_path_raw, "SWP/")
path_weight     <- paste0(data_path_raw, "weight/")
```

Define a custom theme for plots

```{r, echo=FALSE, results='hide'}
my_theme <- ggthemes::theme_few() +
  theme(
    text = element_text(size = 11),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 11),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 12, hjust = 0.5),
    legend.position = "none"
  )
```

### Load Metadata

Load metadata about the experiment (Plant_IDs, treatments, etc).

NB : I created a Plants_ID file to be consistent with the design of the project.

```{r}
# Metadata <- read.csv(file = paste0(data_path_clean, "metadata.csv"))

# Plants_ID <- Plants %>% 
#   mutate(Plant_ID = Plant) 
# Plants_ID <- Plants_ID[c("Plant_ID", "PSY", "Teros", "Treatment")]
# write.csv(x = Plants_ID, file = paste0(data_path_clean, "Plants_ID.csv"), row.names = F)

Plants_ID <- read.csv(paste0(data_path_clean, "Plants_ID.csv"))
```

## 1) Load LWP data

### Extract from raw files

Leaf Water Potential data from psychrometers are stored in separated `.csv` files.

We use the custom function `load_LWP` to load and process each csv file in the provided path.

The result is a single table containing Date, Time, LWP value and the ID of plants.

The single file is writen as `LWP_raw.csv` in the raw folder.

```{r, eval=FALSE}
# Compile all PSY files into one dataframe
LWP_raw <- load_LWP(path = path_LWP)

# Change the PSY id by Plant_ID
LWP_raw <- left_join(LWP_raw, Plants_ID[c("Plant_ID", "PSY")]) %>%
  select(-c("PSY"))

head(LWP_raw)

# Write to raw folder
write.csv(x = LWP_raw, file = paste0(data_path_raw, "LWP_raw.csv"), row.names = F)
```

### Load compiled raw file

Once the raw files are compiled, we can simply load the processed csv file.

```{r, message=FALSE, warning=FALSE}
# Load from Raw folder
LWP_raw <- read.csv(paste0(data_path_raw, "LWP_raw.csv")) %>% 
  mutate(DateTime = anytime(DateTime)) %>% 
  mutate(DateTime = as.POSIXct(DateTime, format = "%Y-%m-%d %H:%M:%S"))

p1 <- ggplot(left_join(LWP_raw, Plants_ID, by = "Plant_ID")) + 
  
  geom_line(aes(x = DateTime, 
                y = LWP, 
                color = Treatment,
                group = Plant_ID)) +
  
  ylim(-1.5, 0) +
  
  ylab("Leaf Water Potantial [MPa]") +
  
  facet_wrap(~Plant_ID, 
             # scales = "free_y",
             ncol = 2) +
  
  theme_few() + theme(axis.text.x = element_text(angle = 90)) 
  # my_theme +
  # theme(strip.text.x = element_blank())
  # ggtitle("Clean LWP")

p1
ggsave(filename = "LWP_raw.png", plot = p1, device = "png", path = plot_path, width = 9, height = 4)
```

### Clean raw file with cleaning algorithm and write clean file.

We see that data is messy, so we implemented an algorithm to clean LWP data. It consists of creating buffers per day, and for each buffer we run a bunch of tests to check if we want to keep the buffer of not.

-   `check.na` evaluates how many NAN are in the buffer. If there is more NAN than the `nan.th` proportion, the buffer is discarded.

-   `check.zero` evaluates how many zeros are in the buffer. If there is more zeros than the `zero.th` proportion, the buffer is discarded.

-   `check.cycles` evaluates if LWP at night is lesser than LWP at day. If the mean LWP at day is higher that the mean LWP at night multiplied by `cycle.alpha`, the buffer is discarded.

-   `check.slope` evaluates the mean slope of the buffer via a linear model. If the absolute value of the slope is higher than `slope.th`, the buffer is discarded.

-   `check.level` evaluates the statistical mean of the buffer compared to the mean of the other buffers. If \$\bar{µ}*{LWP, i} \<* \bar{µ}{LWP} - \sigma*{LWP}\alpha \$ or \$\bar{µ}{LWP, i} \> \bar{µ}{LWP} +\* \sigma{LWP}\alpha \$, the buffer is discarded.

Finally, the clean dataframe is written to storage.

```{r, message=FALSE, warning=FALSE}
source("../src/LWP_algorithm.R")

LWP_clean <- clean.LWP(LWP_data = LWP_raw,
                       check.na     = T,
                       check.zero   = T, 
                       check.cycles = T, 
                       check.slope  = F, 
                       check.level  = F, 
                       nan.th       = 0.5, 
                       zero.th      = 0.5, 
                       cycle.alpha  = 0.5, 
                       slope.th     = 0.000002, 
                       level.alpha  = 2)

LWP_clean <- LWP_clean[c("Date", "Time", "DateTime", "LWP", "Plant_ID")]

# Additionnal cleaning
LWP_clean <- LWP_clean %>%
  filter(LWP > -3)

# Write to storage
write.csv(x = LWP_clean, file = paste0(data_path_clean, "LWP_clean.csv"), row.names = F)
```

### Load clean file

```{r, message=FALSE, warning=FALSE}
LWP_clean <- read.csv(paste0(data_path_clean, "LWP_clean.csv")) %>% 
  mutate(DateTime = anytime(DateTime)) %>% 
  mutate(DateTime = as.POSIXct(DateTime, format = "%Y-%m-%d %H:%M:%S"))

# Add NA so that there is gaps in geom_line
temp <- expand.grid(DateTime = unique(LWP_clean$DateTime),
                    Plant_ID = unique(LWP_clean$Plant_ID)
                    )

LWP_clean <- left_join(temp, LWP_clean, by = c("Plant_ID", "DateTime"))

p2 <- ggplot(left_join(LWP_clean, Plants_ID, by = "Plant_ID")) + 
  
  geom_line(aes(x = DateTime, 
                y = LWP, 
                color = Treatment,
                group = Plant_ID)) +
  
  # ylim(-3, 0) +
  
  ylab("Leaf Water Potantial [MPa]") +
  
  facet_wrap(~Plant_ID, 
             # scales = "free_y",
             ncol = 2) +
  
  theme_few() + theme(axis.text.x = element_text(angle = 90)) 
  # my_theme +
  # theme(strip.text.x = element_blank())
  # ggtitle("Clean LWP")

p2
ggsave(filename = "LWP_clean.png", plot = p2, device = "png", path = plot_path, width = 9, height = 4)
```

We can make a quick graph of the comparison between Raw and Clean LWP

```{r, message=FALSE, warning=FALSE}
p <- ggplot() +
  geom_line(data = LWP_raw, aes(x = DateTime, y = LWP), color = "red") +
  geom_line(data = LWP_clean, aes(x = DateTime, y = LWP), color = "green") +
  # ylim(-3, 0.5) +
  facet_wrap(~Plant_ID) +
  theme_bw()

ggsave(filename = "LWP_comparison.png", 
       plot = p, device = "png", path = plot_path, width = 10, height = 4)
```

### Investigate Day and Night cycles

```{r, message=FALSE, warning=FALSE}
data2plot <- left_join(LWP_raw, Plants_ID, by = "Plant_ID") %>% 
  mutate(Hours = hour(DateTime))

dnc_plot <- ggplot(data2plot, aes(x = as.factor(Hours), y = LWP)) +
  # geom_violin(aes(x = as.factor(Hours), y = LWP), fill = "white") +
  geom_jitter(alpha= 0.5) +
  stat_summary(fun.y = median, geom = "point", color = "red", size = 3) +
  facet_wrap(~Plant_ID) + theme_few()

dnc_plot

ggsave(filename = "DNC.png", plot = dnc_plot, device = "png", path = plot_path, width = 10, height = 10)
```


## 2) Load SWP data

### Extract from sensor files and compile into a single raw file

Soil Water Potential data is stored in a `.DAT` file.

We coded a little function to convert that data into a dataframe.

```{r, eval=FALSE, message=FALSE, warning=FALSE}
# Extract from DAT files
SWP_raw <- load_SWP(path_SWP = path_SWP)

# Replace Teros_ID by Plant_ID and set DateTime as POSIXct
SWP_raw <- left_join(SWP_raw, Plants_ID[c("Plant_ID", "Teros")], by = c("Teros_ID" = "Teros")) %>% 
    select(c("DateTime", "SWP", "Plant_ID")) %>% 
    mutate(DateTime = as.POSIXct(DateTime, format = "%d/%m/%Y %H:%M"))

# Write to disk
write.csv(x = SWP_raw, file = paste0(data_path_raw, "SWP_raw.csv"), row.names = F)
```

### Load raw file, clean it a bit and write to a clean file

```{r, message=FALSE, warning=FALSE}
# Load SWP_raw
SWP_raw <- na.omit(read.csv(paste0(data_path_raw, "SWP_raw.csv"))) %>% 
  mutate(DateTime = anytime(DateTime)) %>% 
  mutate(DateTime = as.POSIXct(DateTime, format = "%Y-%m-%d %H:%M:%S"))

# Filter data so DateTime match with LWP
SWP_clean <- SWP_raw %>% 
  filter(DateTime >= min(LWP_clean$DateTime) & DateTime <= max(LWP_clean$DateTime)) %>% 
  filter(DateTime > as.POSIXct("2024-10-25"))

# Write to disk
write.csv(x = SWP_clean, file = paste0(data_path_clean, "SWP_clean.csv"), row.names = F)
```

```{r, message=FALSE, warning=FALSE}
SWP_plot <- ggplot(left_join(SWP_clean, Plants_ID, by = "Plant_ID")) +
  geom_line(aes(x =  DateTime, y = SWP, color = Treatment)) +
  facet_wrap(~Plant_ID) +
  theme_bw() + my_theme

SWP_plot

ggsave(filename = "SWP_clean.png", 
       plot = SWP_plot, device = "png", path = plot_path, width = 9, height = 4)
```

## 3) Weight

### Extract from sensor output files and compile into one single raw file

```{r, eval=FALSE, message=FALSE, warning=FALSE}
# Load raw data
W_raw <- load_Weigth(path = path_weight, skip = 1) %>% 
  inner_join(Plants_ID[c("Plant_ID", "Scale")], by = "Scale") %>% 
  mutate(Date = as.Date(Date, format = "%d/%m/%Y")) %>%
  select(-c("Scale"))

# Write to disk
write.csv(x = W_raw, file = paste0(data_path_raw, "W_raw.csv"), row.names = F)
```

### Load raw file, clean a bit and write clean file

```{r, message=FALSE, warning=FALSE}
# Load raw file
W_raw <- unique(read.csv(paste0(data_path_raw, "W_raw.csv"))) %>% 
  mutate(DateTime = anytime(DateTime)) %>% 
  mutate(DateTime = as.POSIXct(DateTime, format = "%Y-%m-%d %H:%M:%S"))

# Cleaning
W_clean <- W_raw %>% 
  filter(DateTime >= as.POSIXct("2024-10-24 11:00:00"))

# Check plot
W_plot <- ggplot(W_clean %>% left_join(Plants_ID, by = "Plant_ID")) +
  geom_line(aes(x = DateTime, y = Weight, color = Treatment)) +
  facet_wrap(~Plant_ID) +
  theme_bw() + my_theme

W_plot

ggsave(filename = "W_clean.png", 
       plot = W_plot, device = "png", path = plot_path, width = 9, height = 4)

# Write to disk
write.csv(x = W_clean, file = paste0(data_path_clean, "W_clean.csv"), row.names = F)
```

## 4) Compute Transpiration rate

We use a custom function to compute transpiration rate $T_r$ from weight time series using the formula :

$$ 
T_{r, t} [g \cdot h^{-1}] = (W_{t-1} - W_{t}) / \Delta t  
$$

```{r, message=FALSE, warning=FALSE}
# Load clean Weight file
W_clean <- unique(read.csv(paste0(data_path_clean, "W_clean.csv"))) %>% 
  mutate(DateTime = anytime(DateTime)) %>% 
  mutate(DateTime = as.POSIXct(DateTime, format = "%Y-%m-%d %H:%M:%S"))

# Compute Tr using custom function
Tr_raw <- compute_Tr(W = W_clean, th = 20, DT.format = "%Y-%m-%d %H:%M:%S")

# Write to disk
write.csv(x = Tr_raw, file = paste0(data_path_raw, "Tr_raw.csv"), row.names = F)
```

### Load Tr raw from file

```{r, message=FALSE, warning=FALSE}
# Read Tr raw fil from folder
Tr_raw <- unique(read.csv(paste0(data_path_raw, "Tr_raw.csv"))) %>% 
  mutate(DateTime = anytime(DateTime)) %>% 
  mutate(DateTime = as.POSIXct(DateTime, format = "%Y-%m-%d %H:%M:%S"))

Tr_plot <- ggplot(left_join(Tr_raw, Plants_ID, by = "Plant_ID")) +
  geom_line(aes(x = DateTime, y = Tr, color = Treatment)) +
  facet_wrap(~Plant_ID) +
  ylab("Tr [g/h]") +
  theme_bw() + my_theme

Tr_plot

ggsave(filename = "Tr_raw.png", 
       plot = Tr_plot, device = "png", path = plot_path, width = 9, height = 4)
```
